\chapter{Source code of programs used in experiments}\label{appendixa}

\section{Single source shortest paths}
\begin{Verbatim}[label=Shortest paths - SparkDatalog]
val query = """ declare Path(int v, int dist aggregate Min).
                 Path(x, d) :- s == """ + sourceId + """ , Edge(s, x, d).
                 Path(x, d) :- Path(y, da), Edge(y, x, db), d = da + db."""
val resultDatabase = database.datalog(query)
\end{Verbatim}

\vspace{0cm}

\begin{Verbatim}[label=Shortest paths - Spark]
val initialGraph = graph.mapVertices((id, _) => if (id == sourceId) 0.0 else Double.PositiveInfinity)
val sssp = initialGraph.pregel(Double.PositiveInfinity)(
    (id, dist, newDist) => math.min(dist, newDist), // Vertex Program
    triplet => \{  // Send Message
      if (triplet.srcAttr + triplet.attr < triplet.dstAttr) \{
        Iterator((triplet.dstId, triplet.srcAttr + triplet.attr))
      \} else \{
        Iterator.empty
      \}
    \},
    (a,b) => math.min(a,b)) // Merge Message
val shortestPaths = sssp.vertices.map(v => (v._1, v._2))
\end{Verbatim}


\section{Triangles counting}
\begin{Verbatim}[label=Triangles counting - SparkDatalog]
val query = """ |declare Triangle(int v, int w, int u).
                |declare Total(int a, int b aggregate Sum).
                |Triangle(x, y, z) :- Edge(x, y), x < y, Edge (y, z), y < z, Edge(x, z).
                |Total(a, c) :- Triangle(x, y, z), a = 1, c = 1. """.stripMargin
val resultDatabase: Database = database.datalog(query)
\end{Verbatim}

\vspace{0cm}

\begin{Verbatim}[label=Triangles counting - Spark]
val canonicalEdges = edgesRdd.filter(Function.tupled(_ < _)).distinct().cache()
val swappedEdgesRdd = canonicalEdges.map(_.swap)
val pathOf2 = swappedEdgesRdd.join(canonicalEdges).map( { case (y, (x, z)) => (x, (y, z)) } )
val triangle = pathOf2.join(canonicalEdges)
  .filter({ case (x, ((y, z), zp)) => z == zp })
  .map({ case (x, ((y, z), zp)) => (x, y, z) })
val count = triangle.count()
\end{Verbatim}


\section{Connected components}
\begin{Verbatim}[label=Connected components - SparkDatalog]
val query = """ |declare Component(int n, int component aggregate Min).
                |declare ComponentId(int n).
                |Component(n, i) :- Node(n), i = n.
                |Component(n, i) :- Component(p, i), Edge(p, n).
                |ComponentId(id) :- Component(x, id). """.stripMargin
val resultDatabase = database.datalog(query)
\end{Verbatim}

\vspace{0.4cm}

\begin{Verbatim}[label=Connected components - Spark]
val initialGraph = graph.mapVertices((id, _) => id.toInt)
val connectedComponents = initialGraph.pregel(Int.MaxValue)(
  (id, cmp, newCmp) => math.min(cmp, newCmp), // Vertex Program
  triplet => \{  // Send Message
    if (triplet.srcAttr < triplet.dstAttr) \{
      Iterator((triplet.dstId, triplet.srcAttr))
    \} else \{
      Iterator.empty
    \}
  \},
  (a, b) => math.min(a, b)) // Merge Message
\end{Verbatim}


